Pangenome graph layout by Path-Guided Stochastic Gradient Descent

### 2024-02-19 manuscript/main.tex
feedback Niklas
supplementary

	\red{The classical force model of state of the art generic graph algorithms, such as FM³-based ones, places nodes according to their attractive and repulsive forces.
	This force can be seen as equivalent to how our 2D PG-SGD moves the nodes’ ends in 2D: If the nucleotide distance of the randomly chosen path steps is smaller than the layout distance of the nodes' ends, we move them closer together (“attractive force”), else we move them further away (“repulsive force”). 
	However, the key difference here is that this approach is path-guided: Paths represent biological sequences in pangenome graphs, so it is as if PG-SDG considers a “biological force” for placing the graph nodes.
	So theoretically, it would be possible to combine our approach with a force-directed one. 
	Combining both methods, we might get the best of both worlds: Multi-threadable PG-SGD iteratively applied to different graph-layout-levels. 
	We can imagine that such an approach can lead to a further speedup when calculating the layout.
	However, for generic graphs, this would only work if path information for each node could be added: We would replace the classical physical simulation approach with our path-guided method. 
	If such information is not available, one could randomly cover the graph with paths.
	This function is already provided in \textit{odgi cover}. 
	However, this is an NP-hard problem and so far didn't work very well in practice.}

@@

	\clearpage
	\setcounter{page}{1}

	\beginsupplement

	\section{Supplement}

	 \subsection{Supplementary data}
	 %\subsection{Quantization} % DURING REVISION, IF NEEDED
	 %More details about your clever quantization.
	 \subsubsection{Performance evaluation}
	 \label{sec:performance}
	 The results of the performance evaluation are given in Table~\ref{tab:layout}.
	 \input{table_layout}
	 %\FIXME{What about the SMT thingy?} % DURING REVISION, IF NEEDED
	\subsubsection{1D visualizations}	
	The 1D PG-SGD algorithm creates a 1D layout of the nodes of the graph.
	Theoretically, it is possible that 2 nodes have the same 1D coordinate or overlap.
	But, in our 1D visualizations, we arrange the nodes from left to right.
	Therefore, we project the 1D coordinates into a 1D node order: We sort the final layout by graph component, graph position, and node rank.
	\input{fig_1d_sorts}

>>>>

	\red{The classical force model of state of the art generic graph algorithms, such as FM³-based ones, places nodes according to their attractive and repulsive forces.
	This force can be seen as equivalent to how our 2D PG-SGD moves the nodes’ ends in 2D: If the nucleotide distance of the randomly chosen path steps is smaller than the layout distance of the nodes' ends, we move them closer together (“attractive force”), else we move them further away (“repulsive force”). 
	However, the key difference here is that this approach is path-guided: Paths represent biological sequences in pangenome graphs, so it is as if PG-SDG considers a “biological force” for placing the graph nodes.
	So theoretically, it would be possible to combine our approach with a force-directed one. 
	Combining both methods, we might get the best of both worlds: Multi-threadable PG-SGD iteratively applied to different graph-layout-levels. 
	We can imagine that such an approach can lead to a further speedup when calculating the layout.
	However, for generic graphs, this would only work if path information for each node could be added: We would replace the classical physical simulation approach with our path-guided method. 
	If such information is not available, one could randomly cover the graph with paths.
	This function is already provided in \textit{odgi cover}. 
	However, this is an NP-hard problem and our preliminary solutions proved ineffective.}

@@

	%\clearpage
	%\setcounter{page}{1}

	%\beginsupplement

	%\section{Supplement}
	%
	% \subsection{Supplementary data}
	 %\subsection{Quantization} % DURING REVISION, IF NEEDED
	 %More details about your clever quantization.
	% \subsubsection{Performance evaluation}
	% \label{sec:performance}
	% The results of the performance evaluation are given in Table~\ref{tab:layout}.
	% \input{table_layout}
	 %\FIXME{What about the SMT thingy?} % DURING REVISION, IF NEEDED
	%\subsubsection{1D visualizations}	
	%The 1D PG-SGD algorithm creates a 1D layout of the nodes of the graph.
	%Theoretically, it is possible that 2 nodes have the same 1D coordinate or overlap.
	%But, in our 1D visualizations, we arrange the nodes from left to right.
	%Therefore, we project the 1D coordinates into a 1D node order: We sort the final layout by graph component, graph position, and node rank.
	%\input{fig_1d_sorts}
